Articles related to chronological mutation order have been retrieved mainly from Google Scholar \footnote{\url{https://scholar.google.com}}, based on the keywords picked from this work´s title: \emph{mutation order relevance} and \emph{chronological gene alterations}. The rest of the literature has been derived from the first results found on the web if these were providing sufficient insight on the matter.
Introduction and staring perspective are included in all the background information provided prior to starting this project.
\\

Methods of dysregulated path discovery and other approaches have been collected from starting papers references or student’s own knowledge acquired from professional career.
\\

First, essential cancer-related works have been checked, from which difficulties like treatment unresponsiveness and divergent patient evolution were collected, both from conclusions and discussions.

Then, focusing on this work's proposal, gene order chronology papers were retrieved and scanned for supportive examples. Amongst the examined references, pathway dysrgulation appeared as a promising line of study, which ended up as a whole section due its relation to our theme and relative importance in the new advances in terms of research.

Along the full search, complications related to chronology and heterogeneity seemed to have multiple sources. Amongst the major ones, phylogenetic tree retrieval and timeline establishment were highlighted.

Last, combining biological knowledge and its implications in the order-based approach, we included germline analysis. This point is important because inheritance challenges the full order-based scheme discussed in this work.